pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-02-26 article.tex
Update scripts

There are plenty of diversification models: 
constant-birth [Yule, 19..],
birth-death [\cite{nee1994reconstructed}], 
time-dependent [Rabosky D., LASER package?], 
diversity dependent [\cite{etienne2011diversity}], 
protracted birth-death [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
multiple-birth death [Laudanno, Bilderbeek, Haegeman \& Etienne, in preparation],
a rate shifting birth-death model [the SLS model. Laudanno, Haegeman \& Etienne, in preparation]
punctuated equilibrium (as known as "early burst") [\cite{harmon2010early}],
dependent on a binary trait [\cite{maddison2007estimating}], 
multiple state trait [\cite{fitzjohn2012diversitree}],
concealed state [\cite{beaulieu2016detecting}] 
or even a combination of multiple concealed and 
observable states [\cite{herrera2018detecting}]. 
\giovanni{we might need to polish this list a little bit}

@@

\iffalse
One of the purposes of modelling a complex systems
is to hypothesize the main driving factors of a process 
and trying to prove that they are indeed able 
to explain the main features of the phenomenon you want to explain.
Due to this, there are diminishing returns on adding complexity
to a model: the job of a modeller is to try to
disentangle the intrinsic complexity of a system 
highlighting the main reasons of its behaviour. 
In such a light you don't want your model to be unnecessary complex.
Additionally, developing a likelihood model takes time, 
human resources, fundings and can strain computational resources.
\fi

@@

\iffalse
When assuming that a new diversification model is worth the
(philosophical) effort, there are many ways to measure
if the new model indeed gives a better explanation. 
\fi

@@

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls the popular Bayesian inference tool 
BEAST2 [\cite{bouckaert2014beast}]. 
With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to validate new diversification models.
Such a new diversification model can then be added to BEAST2,
which already supplies multiple speciation models (among others, 
Yule and birth-death) and also allows to add 
new diversification models via the BEAST2 package system.
\richel{rework}

@@

\begin{sidewaysfigure}
  \centering
  \includegraphics[width=\textwidth]{overview.png}
  \caption{pirouette pipeline}
  \label{fig:pipeline}
\end{sidewaysfigure}

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given reconstructed phylogeny 
\giovanni{
  We should mention here that the phylogenies that are meant to be 
  submitted to pirouette are simulated phylogenies generated according 
  to the model we want to test
}. 
To be able to explain how we do this, we first give definitions
for generative model and inference model: 
a generative model is a combination of site model, clock model and tree prior,
used to simulate alignments. An inference model is a combination of 
site model, clock model, tree prior, an optional node calibration prior
and setup of the Markov chain Monte Carlo (MCMC) algorithm, 
used to produce a posterior. Node calibration priors, called 'most
recent common ancestor' (MRCA) priors within BEAST2 allow to create a 
time-calibrated phylogeny, by specifying a normal distribution around
an expected time when two or more taxa's 
\giovanni{
  I have the feeling that taxa should be plural here. 
  Should we mention that the group must be monophyletic?
} ancestors diversified.

\subsection{pirouette's pipeline}
The pipeline is summarized by the following steps, which will be described in detail below:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item from the alignment an inference model is chosen;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny to estimate the error we make. The higher the match, the lower the error;
\end{enumerate}
\richel{figure to pipeline here}
There is also the option to generate a 'twin tree',
that goes through the same pipeline.

@@

\giovanni{We need to discuss the results we got from the 3 experiments.}
\richel{Only if the journal requires this}

>>>>

There are plenty of diversification models: 
birth-death models that have rates that are 
constant [\cite{nee1994reconstructed}], 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
diversity-dependent [\cite{etienne2011diversity}],
or can shift [the SLS model. Laudanno, Haegeman \& Etienne, in preparation].
There are birth death models in which speciation is a process that 
takes time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
can co-occur [Laudanno, Bilderbeek, Haegeman \& Etienne, in preparation],
dependent on a trait that is binary [\cite{maddison2007estimating}], 
has multiple states [\cite{fitzjohn2012diversitree}],
has a concealed state [\cite{beaulieu2016detecting}] 
or has multiple observed and concealed states [\cite{herrera2018detecting}].

@@

@@

@@

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls the popular Bayesian inference tool 
BEAST2 [\cite{bouckaert2014beast}]. 
With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to judge the necessity of a diversification 
model.
Upon finding a diversification model is warranted, 
it can then be added to BEAST2's existing set of (birth-death and 
coalescent) models via the BEAST2 package system.

@@

\begin{sidewaysfigure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{pirouette pipeline}
  \label{fig:pipeline}
\end{sidewaysfigure}

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given reconstructed phylogeny, within the context of
a simulation study on a (possibly new) diversification model. Such
a study will typically construct phylogenies for different combinations of
the diversification model's parameters, to assess under which scenario's
the error BEAST2 makes cannot be ignored anymore. Within such a study,
there need be many replicates.

To be able to explain how \verb;pirouette; measures the inference error
BEAST2 makes, we first give two definitions of what
we mean by a generative model and by an inference model: 
a generative model is a combination of site model, clock model and tree prior,
used to simulate alignments. An inference model is a combination of 
site model, clock model, tree prior, an optional node calibration prior
and setup of the Markov chain Monte Carlo (MCMC) algorithm, 
used to produce a posterior. Node calibration priors, called 'most
recent common ancestor' (MRCA) priors within BEAST2 allow to create a 
time-calibrated phylogeny, by specifying a normal distribution around
an expected time when the ancestors of two or more taxa diversified.

\subsection{pirouette's pipeline}
The pipeline is summarized by the following steps,
which will be described in detail below:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item from the alignment an inference model is chosen;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny to estimate the error we make. The higher the match, the lower the error;
\end{enumerate}
The pipeline is visualized in \ref{fig:pipeline}.
There is also the option to generate a 'twin tree',
that goes through the same pipeline.

@@

In figure ...
